Understanding Implications of Dataset Choice for Feature Effect Estimation: A Simulation-Based Investigation through Error Decomposition

### 2025-02-11 paper/paper.tex
paper: add chapter and appendix on mc error ablation study results

@@

@@

\clearpage
\section{Further Investigation of Main Experiment Results}\label{app:main-experiment-further-investigation}

@@

>>>>

Finally, we also observe that the Monte Carlo variance --- as expected --- reduced with larger dataset size.
We investigate this effect further in the next ablation study.

@@

We show the effect of the dataset size on the Monte Carlo variance for the features of the Friedman1 dataset
for PDP and ALE in \textbf{Fig.\@~\ref{fig:mc-variance}}. The full results are provided in \textsc{Appendix~\ref{app:additional-mc-results}}.
Naturally, we observe that the Monte Carlo variance decreases with increasing dataset size (except for $x_3$ and $x_4$ for PDP,
where the variance is generally very small). We also see differences between the features: for more complex features like $x_1$ and $x_2$,
which include interactions, the Monte Carlo variance is generally higher compared to simpler features like $x_3$ and $x_4$.
Nonetheless, we observe an interesting difference between PDP and ALE. While they converge to a similar Monte Carlo variance for  % chktex 13
large dataset sizes, the initial Monte Carlo variance for very small sample sizes is substantially higher for ALE than for PDP.  % chktex 13
This demonstrates that ALE is more sensitive to the dataset size than PDP. This is most likely due to the fact that ALE  % chktex 13
estimates effects per interval and thus requires more samples to estimate the effect accurately compared to PDP, which
uses the full dataset to estimate the effect. Also note that this analysis shows that the Monte Carlo variance introduced
by estimating the ground truth feature effects in the main study is for most cases negligible compared to the other error sources.

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/mean_mc_variance_friedman1_pdp.png}
        \caption{Mean Monte Carlo variance for PDP}
        \label{fig:mc-variance-pdp}  % chktex 24
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/mean_mc_variance_friedman1_ale.png}
        \caption{Mean Monte Carlo variance for ALE}
        \label{fig:mc-variance-ale}  % chktex 24
    \end{subfigure}
    \caption{Mean Monte Carlo variance for PDP and ALE on Friedman1 dataset.
    For each data size $n_{mc}$, variances are averaged over all grid points.
    Second linear feature ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability.}
    \label{fig:mc-variance}  % chktex 24
\end{figure*}

In addition, a detailed analysis of the Monte Carlo variance per grid point reveals that for PDP,
the variance mostly depends on the sample density, while for ALE, the variance is also influenced by the
shape of the underlying function. We provide detailed results in \textsc{Appendix~\ref{app:additional-mc-results}}.

@@

\clearpage
\section{Further Investigation of Main Experiment Results}\label{app:main-study-further-investigation}

@@

\clearpage
\section{Additional Results for Ablation Study: Datasize Impact on Monte Carlo Variance}\label{app:additional-mc-results}

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/SNC-all/mean_mc_variance_pdp.png}
        \caption{Mean Monte Carlo variance on SimpleNormalCorrelated for PDP}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/SNC-all/mean_mc_variance_ale.png}
        \caption{Mean Monte Carlo variance on SimpleNormalCorrelated for ALE}
    \end{subfigure}
    \caption{Mean Monte Carlo variance for PDP and ALE on SimpleNormalCorrelated dataset.
    For each data size $n_{mc}$, variances are averaged over all grid points.
    Note that $x_3$ and $x_4$ are dummy features.}
    \label{fig:mean-mc-variance-snc}  % chktex 24
\end{figure*}

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/Friedman1-all/mean_mc_variance_pdp.png}
        \caption{Mean Monte Carlo variance on Friedman1 for PDP}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/Friedman1-all/mean_mc_variance_ale.png}
        \caption{Mean Monte Carlo variance on Friedman1 for ALE}
    \end{subfigure}
    \caption{Mean Monte Carlo variance for PDP and ALE on Friedman1 dataset.
    For each data size $n_{mc}$, variances are averaged over all grid points.
    Note that $x_6$ and $x_7$ are dummy features.}
    \label{fig:mean-mc-variance-friedman1}  % chktex 24
\end{figure*}

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/SNC-all/mc_variance_pdp.png}
        \caption{Monte Carlo variance on SimpleNormalCorrelated for PDP}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/SNC-all/mc_variance_ale.png}
        \caption{Monte Carlo variance on SimpleNormalCorrelated for ALE}
    \end{subfigure}
    \caption{Monte Carlo variance for PDP and ALE on SimpleNormalCorrelated dataset.
    The variances for different data sizes $n_{mc}$ (colors) are shown for each grid
    point over the range of each feature (x-axis). The ground truth feature effect is
    shown as dashed line for reference, using a different y-axis scale omitted for
    clarity. Note that $x_3$ and $x_4$ are dummy features.}
    \label{fig:mc-variance-snc}  % chktex 24
\end{figure*}

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/Friedman1-all/mc_variance_pdp.png}
        \caption{Monte Carlo variance on Friedman1 for PDP}
    \end{subfigure}
    \\[10pt]
    \vfill
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{img/Friedman1-all/mc_variance_ale.png}
        \caption{Monte Carlo variance on Friedman1 for ALE}
    \end{subfigure}
    \caption{Monte Carlo variance for PDP and ALE on Friedman1 dataset.
    The variances for different data sizes $n_{mc}$ (colors) are shown for each grid
    point over the range of each feature (x-axis). The ground truth feature effect is
    shown as dashed line for reference, using a different y-axis scale omitted for
    clarity. Note that $x_6$ and $x_7$ are dummy features.}
    \label{fig:mc-variance-friedman1}  % chktex 24
\end{figure*}
